Utilizing Parallel Workers: \\LLVM's Vectorization Plan

### 2024-11-20 paper/paper.tex
paper

There are various ways to transform the control flow of a loop to accommodate operating on widened
vector registers. Figure \ref{fig:inner-loop-vec} shows an example of an if-conversion. Some more
examples can also be found in LLVM's Auto-Vectorization Documentation~\cite{llvmvec}.

@@

Figure \ref{fig:outer-loop-vec} shows a nested loop code snippet where vectorization over the
outer loop would be beneficial.

@@

\subsection{Vectorization Legality}
Not all loops can be vectorized. There are various factors that can completely hinder any
vectorization attempts.

% TODO

\subsection{Vectorization Costs}
While it may just not be legal to vectorize a given control flow, sometimes the vectorized code 
might also just perform worse than its scalar equivalent.

Some control flow transformations that seem beneficial at compile time might reveal a performance
bottleneck at runtime. Considering the code snippet from Figure \ref{fig:inner-loop-vec}, it might
be that at runtime almost all of \texttt{z[i]} evaluate to false. The branch predictor of a CPU
would then be able to guess the correct branch in the scalar code. The vectorized version would
always load the \texttt{y[i]} into memory only for the addition to be masked away. There are
different approaches to avoid such runtime penalties such as branch-on-superword-condition code 
(BOSCC)~\cite{10.5555/1299042.1299055,llvmboscc} or active-lane 
consolidation (ALC)~\cite{10.1007/s11227-022-04359-w,10.5555/3615924.3615932}.

Some vector instructions can be quite costly on most CPUs. An example would be gather/scatter
or shuffeling and permutation instructions. Control flow with a lot of non-adjacent 
memory load/stores or operations that combine different vector elements such as reductions
can decrease vectorization performance.

Additionally the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow oftentimes simply includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle % TODO

>>>>

There are various ways to transform the control flow of a loop to accommodate operating on widened
vector registers. Figure~\ref{fig:inner-loop-vec} shows an example of an if-conversion. Some more
examples can also be found in LLVM's Auto-Vectorization Documentation~\cite{llvmvec}.

@@

Figure~\ref{fig:outer-loop-vec} shows a nested loop code snippet where vectorization over the
outer loop would be beneficial.

@@

\subsection{Vectorization Legality}
Not all loops can be vectorized. There are various factors that can ultimately hinder any
vectorization attempts.

Some of the most common vectorization barriers are data dependencies. As the vectorized version
of a code executes for $N$ (Vectorization Factor) data streams simultaneously, it inherently changes
the order of operations. To guarantee that the semantics of the code do not change, all
operations on these $N$ data streams must, generally, be independent. The simplest case
where this is not provided is when the $i$-th loop iteration depends on any of the $(i-N)$-th 
previous iterations.

While explicit data dependencies are easier to see, even a simple loop, as in 
Figure~\ref{fig:inner-loop-vec}, could carry an implicit dependency in line 3, if
the \texttt{x} and \texttt{y} point to overlapping memory regions (pointer aliasing).

Not all data dependencies directly hinder any vectorization. One exception can 
sometimes be made for reduction idioms, as seen in the inner loop in 
Figure~\ref{fig:outer-loop-vec}, where \texttt{sum} would be a reduction variable.

Additionally, the control flow should be primarily straight-line code. While most if-else 
constructs can be flattened to always execute both branches along with masking, the same is not 
valid for switch-case statements or too complex control flows.

\subsection{Vectorization Costs}
While it may not be legal to vectorize a given control flow, the vectorized code sometimes 
performs worse than its scalar equivalent.

Some control flow transformations that seem beneficial at compile time might reveal a performance
bottleneck at runtime. Considering the code snippet from Figure~\ref{fig:inner-loop-vec}, it might
be that almost all \texttt{z[i]} evaluate to false at runtime. The branch predictor of a CPU
would then be able to guess the correct branch in the scalar code. The vectorized version would
always load the \texttt{y[i]} into memory only for the addition to be masked away. Different 
approaches to avoid such runtime penalties are branch-on-superword-condition code 
(BOSCC)~\cite{10.5555/1299042.1299055,llvmboscc} or active-lane 
consolidation (ALC)~\cite{10.1007/s11227-022-04359-w,10.5555/3615924.3615932}.

Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
and permutation instructions. Control flow with many non-adjacent memory loads/stores or 
operations that combine different vector elements, such as reductions, can decrease vectorization 
performance.

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop is executed.
